---
title: Milestone TWO; Testing a Model for an Automated Real-Time Acuity Monitoring System in the Emergency Department
author:  
  - name: Tommy Flynn
    email: tjflynn@emory.edu
    affiliation: Emory University
    footnote: Corresponding Author
address:
    code: Emory University
    address: Find the GitHub repository at [https://github.com/tommyflynn/Project-Milestone-1.git](https://github.com/tommyflynn/Project-Milestone-1.git)
abstract: Patient acuity in the Emergency Department is triaged at the beginning of the care process using the Emergency Severity Index (ESI). Although the ESI is widely used and accepted as the most valid predictor of ED resource consumption in US hospital EDs, its predictive power has limitations that can negatively impact patient flow and safety. An objective measurement of individual resource consumption, taken in the context of all ED care, would allow nurses and administrators to make better decisions with staffing and assignemnts. The question remains, can patient acuity be continuously measured with an objective scale based? I hypothesize that patient acuity in the context of direct interactions. The purpose of this project is to determine if patient acuity in the ED is correlated to patient eigenvector centrality in the network of all face to face ED interactions. A strong correlation between network position and acuity is expected, and a change in network position will likely represent a change in patient condition, something the ESI is not capable of. 
  

bibliography: rfid_bib.bib
output: rticles::elsevier_article
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
library(tidyverse)
library(knitr)
library(pander)
library(ggplot2)
library(dplyr)
```


Backgroun & Objectives
======================

In an Emergency Deparmtent (ED), care is delivered over a network of face to face human interactions. Patients interact with registration staff, then a triage nurse who may decide to discuss the patient with a provider, the provider may then interact directly with the patient, and so on. In this way, the network grows over time, creating a web of care that may correlate with the amount and quality of care delivered to individual patients. 
- The purpose of this study is to explore associations between the network of interactions that take place in the Emergency Department and individual patient acuity. To study this relationship, I will analyze the following; 1) the frequency and duration of all interactions (patients, providers, nurses, technicians, & administrators) that occur in the ED, and 2) individual patients' medical and demographic characteristics. The network structural characteristics will be assessed in relation to the industry standard acuity measure, the Emergency Severity Index (ESI), and potential confounding variables. Using this data will require specific knowledge of the R statistical packages, network analysis, and data science. See Tables 1-4 for my learning goals with respective action items, timeline, and outcomes. 

```{r objectives tables needing reformatting, echo=FALSE}
column_names <- c("Demonstrate effective use of GitHub Version Control", "Demonstrate working knowledge of R Studio & R Markdown", "Create useful visualizations of data", "Apply appropriate statistical methods", "Interpret and communicate results")
row_names <- c( "Action Items", "Timeline", "Outcome")
github <- c("Use GitHub version control throughout project development", "Ongoing", "Complete record of data management & analysis")
rstudio <- c("All data wrangling & analysis in R Studio and all milestones completed in R Markdown", "Ongoing", "Final Project, Presentation, and Website Completed in R Markdown")
visualize <- c("Apply appropriate visualization tools to analysis results", "April 26 2018", "Appropriate Tables and Graphs in final presentation and manuscript")
statistics <- c("Execute rigorous statistical analysis of the data", "Apeil 26 2018", "Statistical tests are appropriate to the data and research purpose and error is adequately minimized")
results <- c("Recognize and communicate important results", "April 26 2018", "Results discussed in the final project speak to the research question and bridge a gap in the literature")
mygoals <- data.frame(github, rstudio, visualize, statistics, results, row.names = row_names)
pander(mygoals, col.names = column_names, align = 'l')
```

Data
====
For this project, I have chosen to use, with permission from Vicki Hertzberg, the same data that I am using for my dissertation research. Data were collected using a prospective longitudinal observational study design with a random sampling of two shifts per week, one day and one night, over the course of a year, from July 2009 to June 2010, for a total of 104 shifts.[@RN1] This strategy was chosen to minimize sampling bias related to seasonal or weekly fluctuations in census, acuity, and ED staffing changes. The purpose of the original study was to describe contact characteristics between patients and staff in the ED of a busy urban hospital to inform cross-infection control measures. Data were collected using a radio-frequency identification system that triangulated patient and staff (nurses, providers, administrators, and clinical support staff) locations with the ED at Emory University Hospital Midtwon. 

##Data Wrangling:
I have requested the original/raw data, which will require cleaning and organizing to meet the needs of my research aims. Data will be maintained in private repositories in the GitHub version control platform. Patient characteristic data will be evaluated for missing or implausible data with discriptive analyses, and RFID generated networks will be included for statistical analysis if variables of network density, centrality, and a network diversity scale are distributed normally across networks. 

Analysis Plan
=============

##Exploratory Analysis

Descriptive statistics of the network data as well as patient demographic data will be evaluated for asssumptions of normality. The data will be skewed in certain predictable ways due to the observed patient populations. The distribution of study subject demographics will be described in tabular format, noting irregularities and potential sources of error.
```{r acuity histogram}
patients <- patients %>%
  filter(Acuity != "Not Recorded")
ggplot(patients, aes(Acuity)) + geom_bar(aes(fill = Acuity))
```
```{r participant type histogram} 
#???????????????????????????????????????????????????????????????????????????????????????????????????
#try to add race as a stack to patients bar??
colors <- c("blue","red","lightskyblue","pink")
ggplot(nodes, aes(participant_type, color = participant_type)) + geom_bar(aes(fill = colors)) + scale_fill_identity("Color legend", guide = "participant_type", labels = c("Patient","RN","MD","Staff")) + scale_color_identity("Color legend", guide = "participant_type", labels = c("Patient","RN","MD","Staff")) + labs(title = "Participant Type Distribution", x = "Participant Type", y = "Count") + sec_axis(race?)
#???????????????????????????????????????????????????????????????????????????????????????????????????
```

##Variables used for final analysis:
__Network Variables__
>  - *Patient eigenvector centrality* (dependant varialbe of interest)
>  - Network density
>  - Network clustering coefficient
>  - Network diversity scale??

__Staff variables__
>  - Title (RN, MD, staff)

__Patient variables__
>  - *Acuity* (ESI, independant variable of interest)
>  - Commorbidities (index)
>  - Gender
>  - Age
>  - Race
>  - Ethnicity
>  - Arrival mode (ambulance v. walk-in)
>  - Education (if available)
>  - Disposition (admission v. discharge)
>  - Length of stay (common measure of quality in the literature used for comparison)
>  - Time before first provider contact (common measure of quality in the literature used for comparison))
  
##Analysis

Muliple linear regression will be used for the final analysis to assess the correlation between patient acuity and patient centrality. Relationships will be evaluated visually (see below) as well as statistically to an alpha of 0.05.
```{r boxplot of mydata, echo=FALSE}
#Set the dataframe up for a scatter of pt acuity/eigenvector centrality
mydata <- Nodes1019 %>% 
  filter(participant_type == "_PAT") %>% 
  select(sid, d8, ID, i, LNLN_WTDEGREE, AGE, Race, arrival_mode, Chief_Complaint, ED_Disposition, Arr_Mode, Sex, MinutesInED, Acuity, acuitycat, BLACKyn, wdegree, Eigenvector_Centrality, Betweenness_Centrality, degree, degree_ps, RELATIVE_DEGREE, Authority)
#scatterplot of eigenvector centrality by acuity

#ggplot(mydata, aes(x = Race, y = MinutesInED, color = Acuity)) + geom_boxplot() + labs(title = "Boxplot of Minutes in the ED by Race, grouped by Acuity", x = "Race", y = "Minutes in ED")
```

```{r boxplot minutes by race w acuity}
ggplot(mydata, aes(y = Eigenvector_Centrality, x = MinutesInED, color = Acuity)) + geom_point() + labs(title = "Boxplot Play", x = "Acuity", y = "Minutes")
```
##Schedule
Milestone 1: February 14th, 2018
-Objectives 
Milestone 2: March 15th, 2018
Final Proposal: April 26th, 2018

Results
=======
Results will be discussed with the visual supplementation of network graphs. This allows the reader to understand concepts that may be difficult to grasp through text alone. 



Discussion
==========
Allocating staff resources in an Emergency Department is an ongoing challenge. How can these results begin to offer solutions to ED staff and patient management? 

What were my primary limitation (both expected and unexpected)?



Conclusion
==========
Did I meet my learning objectives? How would I design a better study next time? 




References {#references.unnumbered} 
==========




