---
title: "Network Analysis Lecture"
author: "Tommy Flynn"
date: "4/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load up package:

```{r}

library(igraph); packageVersion("igraph")

```

#### Making the networks

For small graphs, one easy way to get data into igraph is to open up an editing window, as follows:

```{r}

datafab <- edit(data.frame()) #enter edges

```

You can then turn this data frame into a graph as follows:

```{r}

g <- graph.data.frame(datafab, directed = FALSE) # turn edge list into graph
g # list the graph
summary(g) # find out about the graph
V(g) # list the vertices of the graph
E(g) # list the edges of the graph

```

You can also create a graph from a set of excel files documenting the vertex (or node) names and the edges. First start with an empty graph

```{r}

g <- graph.empty() #start with empty graph
traits <- read.csv("traits.csv", head = FALSE) # read in the vertices
relations <- read.csv("relations.csv", head = FALSE) # read in the edges

# add vertices to the empty graph
g <- add.vertices(g, nv = nrow(traits), name = as.character(traits[,1]), age = traits[,2], gender = as.character(traits[,3]))

# check that graph is being created properly

V(g)$name
V(g)$age
V(g)$gender

# strip out the last names

names <- sapply(strsplit(V(g)$name, " "), "[", 1)
ids <- 1:length(names)
names(ids) <- names

# list the ids

ids

# create an edges list with vertex ids instead of names

from <- as.character(relations[,1])
to <- as.character(relations[,2])
edges <- matrix(c(ids[from], ids[to]), nc = 2)

# list the edges

edges

# add edges to the graph, along with edge attributes

g <- add.edges(g, t(edges), room = as.character(relations[,3]), friend = relations[,4], advice = relations[,5])

```

There are other ways to import pre-existing graphs, provided you know the formats.

As seen above from the Alice and Bob example, we can import an edgelist. Another popular format is that used in the package Pajek. Still another common format is `GML`.

```{r}

# Import an edge list
dolphins <- read.graph("dolphins.txt", format = c("edgelist"))
dolphins

# Leave edges as undirected
dolphins <- read.graph("dolphins.txt", directed=FALSE, format = "edgelist")
dolphins

# Import a graph make in Pajek
karate <- read.graph("karate.txt", format = "pajek")
karate

# Import a graph in GML format

football <- read.graph("football.gml", format="gml")
football 
```

#### Plotting the networks

There are two ways to plot within `igraph`. The function `plot.igraph` gives you a static plot, while the function  `tkplot` gives you more interactive capability. Note you must have XQuartz functioning (Mac users) for `tkplot` to work.

```{r}

# Plotting

plot.igraph(karate)

# Another way

tkplot(karate)

```

You can also do some more with color, as follows:

```{r}

E(g)$color <- "black"
E(g)[room =="Y"]$color <- "red"

tkplot(g, layout = layout.kamada.kawai, edge.color = E(g)$color)

```

#### Graph and Vertex Properties

You can determine properties of the graph as a whole as well as those of individual vertices very easily in `igraph`.

```{r}

# Determine the degree sequence

degree(karate)

# Determine the diameter of the graph

diameter(karate)

# Calculate the closeness centrality for each vertex

closeness(karate)

# Note for directed graphs, you can use the  mode option to determine the in-, out-, or all-closeness properties, as follows:

closeness(karate, mode="in")
closeness(karate, mode = "out")
closeness(karate, mode = "all")

# Calculate the betweenness centrality for each vertex

betweenness(karate)

# And for directed graphs

betweenness(karate, directed = "TRUE")

```


#### Want to try more?

In the repo there are two more datasets. The dataset `lesmis.txt` is a Pajek file denoting the characters in the play "Les Mis", and two characters are related when they appear on stage at the same time. The dataset `adjnoun.gml` is a GML formatted graph that describes the word concurrencies in "David Copperfield".